\Large{Adversarial Training against Systematic Uncertainty}

### 2016-10-20 tex/paper.tex
update toy

If $Z$ takes discrete values, then $p_{\theta_r}$ can be represented e.g. as a
probabilistic classifier $\mathbb{R} \mapsto \mathbb{R}^{|{\cal Z|}}$ whose
output $j$ (for $j=1, \dots, |{\cal Z}|$) is the estimated probability mass
$p_{\theta_r}(z_j|f(X;\theta_f))$. Similarly, if $Z$ takes continuous values and
if we assume some parameteric distribution ${\cal P}$ for $Z|f(X;\theta_f)$
(e.g., a mixture of gaussians over a bounded support, as modeled with a mixture
density network~\citep{bishop1994mixture}), then $p_{\theta_r}$ can be
represented e.g. as network whose output $j$ is the estimated value of the
corresponding parameter $\gamma_j$ of that distribution (e.g., the mean,
variance and mixing coefficients of its components). As in
\citep{nix1994estimating,bishop1994mixture}, the estimated probability density
$p_{\theta_r}(z|f(X;\theta_f))$ can then be evaluated for any $z \in {\cal Z}$.
As further explained in the next section, let us note that the adversary $r$ may
take any form, i.e. it does need to be a neural network, as long as it exposes a
differentiable function $p_{\theta_r}(z|f(X;\theta_f))$ of sufficient capacity
to represent the true distribution within its bounded support.

@@

As a guiding toy example, let us consider the binary classification of 2D
data drawn from multivariate gaussians with equal priors, such that
\begin{align}
    x &\sim {\cal N}((0,0), \begin{bmatrix}
                              1 & -0.5 \\
                              -0.5 & 1
                            \end{bmatrix}) \text{ when } Y=0, \\
    x &\sim {\cal N}((1,1+Z),  \begin{bmatrix}
                              1 & 0 \\
                              0 & 1
                             \end{bmatrix}) \text{ when } Y=1.
\end{align}
The continuous nuisance parameter $Z$ represents in this case our
uncertainty about the exact location of the mean of the second gaussian. Our goal is to
build a classifier $f(\cdot;\theta_f)$ for predicting $Y$ given $X$, but such that
the probability distribution of $f(X;\theta_f)$ is invariant with respect to the
nuisance parameter $Z$.

@@

Let us now consider the adversarial training of $f$ against an adversary $r$
implemented as a mixture density network modeling $Z|f(X;\theta_f)$ as a
 mixture of five gaussians. As for $f$, the network
architecture comprises 2 dense hidden layers of 20 nodes with ReLU activations,
but is followed by an output layer of 15 nodes corresponding to the means,
standard deviations and mixture coefficients of the five gaussians. Output nodes
for the mean values come with linear activations, nodes
for the standard deviations with exponential activations to ensure positivity,
while nodes for the mixture coefficients implement the softmax function to
ensure positivity and normalization. When running
Algorithm~\ref{alg:adversarial-training} as initialized with the classifier $f$
obtained previously, adversarial training effectively reshapes the decision
function so it that becomes almost independent on the nuisance parameter, as
shown in the lower plots of Figure~\ref{fig:toy}. In particular, the conditional
probability densities of the decision scores $f(X;\theta_f)$ are now very
similar to each other, indicating only a small  dependency on the nuisance, as
theoretically expected. The dynamics of adversarial training is illustrated in
Figure~\ref{fig:toy-training}, where the losses ${\cal L}_f$, ${\cal L}_r$ and
${\cal L}_f - \lambda {\cal L}_r$ are evaluated after each iteration of
Algorithm~\ref{alg:adversarial-training}. In the first iterations, we observe
that the global objective ${\cal L}_f - \lambda {\cal L}_r$ is minimized by
making the classifier less accurate, hence the corresponding increase of ${\cal
L}_f$, but which results in a classifier that is more pivotal, hence the
corresponding increase of ${\cal L}_r$ and the total net benefit. As learning
goes, minimizing $E$ then requires making predictions that are more accurate,
hence decreasing ${\cal L}_f$, or that are less dependent on $Z$, hence shaping
$p_{\theta_r}$ so that it becomes more uniform. Indeed, ${\cal L}_f$ eventually
start to slightly decrease, while remaining lower bounded by $\min_{\theta_f}
{\cal L}_f(\theta_f)$ as approximated by the dashed line in the first plot.
Similarly,  ${\cal L}_r$ keeps tending towards the differential entropy $H(Z) =
\log(\sigma \sqrt{2 \pi e}) = 1.419$ of the gaussian distribution with unit variance, as shown by the dashed
line in the second plot. Finally, let us note that the ideal situation of a
classifier that is both optimal and pivotal appears to be unreachable for this
problem, as shown in third plot by the  offset between ${\cal L}_f - \lambda
{\cal L}_r$ and the dashed line approximating $\mathbb{E}_{X} [ H({Y|X}) -
\lambda H(Z)]$.

>>>>

If $Z$ takes discrete values, then $p_{\theta_r}$ can be represented e.g. as a
probabilistic classifier $\mathbb{R} \mapsto \mathbb{R}^{|{\cal Z|}}$ whose
output $j$ (for $j=1, \dots, |{\cal Z}|$) is the estimated probability mass
$p_{\theta_r}(z_j|f(X;\theta_f))$. Similarly, if $Z$ takes continuous values and
if we assume some parameteric distribution ${\cal P}$ for $Z|f(X;\theta_f)$
(e.g., a mixture of gaussians, as modeled with a mixture
density network~\citep{bishop1994mixture}), then $p_{\theta_r}$ can be
represented e.g. as network whose output $j$ is the estimated value of the
corresponding parameter $\gamma_j$ of that distribution (e.g., the mean,
variance and mixing coefficients of its components). As in
\citep{nix1994estimating,bishop1994mixture}, the estimated probability density
$p_{\theta_r}(z|f(X;\theta_f))$ can then be evaluated for any $z \in {\cal Z}$.
As further explained in the next section, let us note that the adversary $r$ may
take any form, i.e. it does need to be a neural network, as long as it exposes a
differentiable function $p_{\theta_r}(z|f(X;\theta_f))$ of sufficient capacity
to represent the true distribution within its bounded support.

@@

As a guiding toy example, let us consider the binary classification of 2D
data drawn from multivariate gaussians with equal priors, such that
\begin{align}
    x &\sim {\cal N}((0,0), \begin{bmatrix}
                              1 & -0.5 \\
                              -0.5 & 1
                            \end{bmatrix}) &\text{ when } Y=0, \\
    x &\sim {\cal N}((1,1+Z),  \begin{bmatrix}
                              1 & 0 \\
                              0 & 1
                             \end{bmatrix}) &\text{ when } Y=1.
\end{align}
The continuous nuisance parameter $Z$ represents in this case our
uncertainty about the exact location of the mean of the second gaussian. Our goal is to
build a classifier $f(\cdot;\theta_f)$ for predicting $Y$ given $X$, but such that
the probability distribution of $f(X;\theta_f)$ is invariant with respect to the
nuisance parameter $Z$.

@@

Let us now consider the adversarial training of $f$ against an adversary $r$
implemented as a mixture density network modeling $Z|f(X;\theta_f)$ as a mixture
of five gaussians. As for $f$, the network architecture comprises 2 dense hidden
layers of 20 nodes with ReLU activations, but is followed by an output layer of
15 nodes corresponding to the means, standard deviations and mixture
coefficients of the five gaussians. Output nodes for the mean values come with
linear activations, nodes for the standard deviations with exponential
activations to ensure positivity, while nodes for the mixture coefficients
implement the softmax function to ensure positivity and normalization. When
running Algorithm~\ref{alg:adversarial-training} as initialized with the
classifier $f$ obtained previously, adversarial training effectively reshapes
the decision function so it that becomes almost independent on the nuisance
parameter, as shown in the lower plots of Figure~\ref{fig:toy}. In particular,
the conditional probability densities of the decision scores $f(X;\theta_f)$ are
now very similar to each other, indicating only a small  dependency on the
nuisance, as theoretically expected. The dynamics of adversarial training is
illustrated in Figure~\ref{fig:toy-training}, where the losses ${\cal L}_f$,
${\cal L}_r$ and ${\cal L}_f - \lambda {\cal L}_r$ are evaluated after each
iteration of Algorithm~\ref{alg:adversarial-training}. In the first iterations,
we observe that the global objective ${\cal L}_f - \lambda {\cal L}_r$ is
minimized by making the classifier less accurate, hence the corresponding
increase of ${\cal L}_f$, but which results in a classifier that is more
pivotal, hence the corresponding increase of ${\cal L}_r$ and the total net
benefit. As learning goes, minimizing $E$ then requires making predictions that
are more accurate, hence decreasing ${\cal L}_f$, or that are less dependent on
$Z$, hence shaping $p_{\theta_r}$ towards the prior $p(z)$. Indeed, ${\cal L}_f$
eventually starts to slightly decrease, while remaining lower bounded by
$\min_{\theta_f} {\cal L}_f(\theta_f)$ as approximated by the dashed line in the
first plot. Similarly,  ${\cal L}_r$ tends towards the differential entropy
$H(Z)$ of the prior (where $H(Z) = \log(\sigma \sqrt{2 \pi e}) = 1.419$ in the
case of a gaussian with unit variance), as shown by the dashed line in the
second plot.
% Finally, let us note that the ideal situation of a
% classifier that is both optimal and pivotal appears to be unreachable for this
% problem, as shown in third plot by the  offset between ${\cal L}_f - \lambda
% {\cal L}_r$ and the dashed line approximating $\mathbb{E}_{X} [ H({Y|X}) -
% \lambda H(Z)]$.
